Model Adaptivity for Goal-Oriented Inference using Adjoints

### 2017-10-21 paper/submitted/round3_png/mathematical_formulation_v6_cmame.tex
formalization of compatibility

Assume the high- and low-fidelity models are \emph{compatible}, in that $U_{LF}\subseteq U_{HF}$ and $Q_{LF}\subseteq Q_{HF}$. Then this compatibility allows one to combine the two models to form a mixed-fidelity model and corresponding mixed-fidelity inverse problem. The low-fidelity model can be used in part of the computational domain, and the high fidelity in the remainder; we denote the mixed-fidelity model and its variables with the subscript $_{MF}$. How one chooses to combine the low- and high-fidelity models can be informed by what aspect of the inferred parameters is of interest.

@@

%
%------------------------------------------------------------------------------%
\subsection[Error Estimate for a Goal-Oriented Inverse Problem]{Error Estimate for a Goal-Oriented Inverse Problem}  \label{sec:deriv}
%------------------------------------------------------------------------------%
%
For a given hierarchy of compatible models, we take the QoI calculated from inferring the parameters with the highest-fidelity model to be the `exact' or `true' solution; we now derive an a posteriori estimate for the error in the QoI from inferring the parameters with a lower-fidelity model.
%
\begin{proposition}
\label{thm:error_estimate}
Consider the high- and low-fidelity inverse problems described by the constrained optimization problems in \Cref{eq:invOptHF,eq:invOptLF}. Let the high- and low-fidelity models be compatible ($U_{LF}\subseteq U_{HF}$ and $Q_{LF}\subseteq Q_{HF}$), and let the forms $a_{HF/LF}:U_{HF/LF} \times Q_{HF/LF} \times U_{HF/LF} \to \Reals$ be three times continuously differentiable with respect to the state $u$ and parameters $q$. Let the observation operator $C:U_{HF/LF}\to\Reals^{n_d}$ be three times continuously differentiable with respect to the state $u$. Also, let the regularization operator $R:Q_{HF/LF}\to\Reals$ be differentiable with respect to the parameter $q$, and the QoI functional $I:Q_{HF/LF}\times U_{HF/LF}\to\Reals$ be differentiable with respect to the state $u$ and parameter $q$.

@@

Extending the property in \Cref{eq:MeqI} to the augmented Lagrangians for the high- and low-fidelity models, we have,
%
\begin{multline}
\label{eq:repIwithM}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})\textrm{.}
\end{multline}
%
Following~\cite{BecVex05}, we rewrite the term $\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})$ using the Fundamental Theorem of Calculus,
%
\begin{equation}
\label{eq:Mantiderivative}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \int\limits_{0}^{1} \mathcal{M}'_{HF}\left(\Psi_{HF} + se\right)\left(e\right) \, \textrm{d}s.
\end{equation}
%
Applying the trapezoidal rule to the integral in~\Cref{eq:Mantiderivative} gives,
%
\begin{equation}
\label{eq:Mprimetrapezoid}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{HF})(e) + \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(e) + \mathcal{R}\textrm{.}
\end{equation}
%
where the remainder $\mathcal{R}$ is given as,
%
\begin{equation}
\label{eq:trapezoidremainder}
\mathcal{R} = \frac{1}{2} \int\limits_{0}^{1} \mathcal{M}'''_{HF}\left(\Psi_{LF} + se\right)(e,e,e) \, s(s-1) \, \textrm{d}s,
\end{equation}
%
which is finite due to the third-order differentiability assumptions. We further note that $\mathcal{M}'_{HF}(\Psi_{HF})(e)$ vanishes since $\Psi_{HF}$ is a stationary point of $\mathcal{M}_{HF}$, and~\Cref{eq:Mprimetrapezoid} can be simplified to,
%
\begin{equation}
\label{eq:beckvex}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{R}\textrm{.}
\end{equation}
%
Combining \Cref{eq:repIwithM} and \Cref{eq:beckvex} we obtain,
%
\begin{multline}
\label{eq:preadj}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}\textrm{,}
\end{multline}
which completes the proof.
\end{proof}
%

>>>>

Assume the high- and low-fidelity models are compatible (formally defined in \Cref{defn:comp}), so that one can combine the two models to form a mixed-fidelity model and corresponding mixed-fidelity inverse problem. The low-fidelity model can be used in part of the computational domain, and the high fidelity in the remainder; we denote the mixed-fidelity model and its variables with the subscript $_{MF}$. How one chooses to combine the low- and high-fidelity models can be informed by what aspect of the inferred parameters is of interest.

@@

%
%------------------------------------------------------------------------------%
\subsection[Error Estimate for a Goal-Oriented Inverse Problem]{Error Estimate for a Goal-Oriented Inverse Problem}  \label{sec:deriv}
%------------------------------------------------------------------------------%
%
For a given hierarchy of compatible models, we take the QoI calculated from inferring the parameters with the highest-fidelity model to be the `exact' or `true' solution. Let compatibility be formally defined as follows:
%
\begin{definition}
The high- and low-fidelity models described by \Cref{eq:invOpt_consHF} and \Cref{eq:invOpt_consLF}, respectively, are \emph{compatible} if $U_{LF}\subseteq U_{HF}$ and $Q_{LF}\subseteq Q_{HF}$. Compatibility allows one to evaluate residuals associated with the high-fidelity inverse problem, such as optimality conditions, at solutions to the low-fidelity inverse problem.
\label{defn:comp}
\end{definition}
% 
We now derive an a posteriori estimate for the error in the QoI from inferring the parameters with a lower-fidelity model.
%
\begin{proposition}
\label{thm:error_estimate}
Consider the high- and low-fidelity inverse problems described by the constrained optimization problems in \Cref{eq:invOptHF,eq:invOptLF}. Let the high- and low-fidelity models be compatible (in the sense of \Cref{defn:comp}), and let the forms $a_{HF/LF}:U_{HF/LF} \times Q_{HF/LF} \times U_{HF/LF} \to \Reals$ be three times continuously differentiable with respect to the state $u$ and parameters $q$. Let the observation operator $C:U_{HF/LF}\to\Reals^{n_d}$ be three times continuously differentiable with respect to the state $u$. Also, let the regularization operator $R:Q_{HF/LF}\to\Reals$ be differentiable with respect to the parameter $q$, and the QoI functional $I:Q_{HF/LF}\times U_{HF/LF}\to\Reals$ be differentiable with respect to the state $u$ and parameter $q$.

@@

Extending the property in \Cref{eq:MeqI} to the augmented Lagrangians for the high- and low-fidelity models, we have,
%
\begin{multline}
\label{eq:repIwithM}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})\textrm{.}
\end{multline}
%
Note that $\mathcal{M}_{HF}(\Psi_{LF})$ is well-defined due to model compatibility. Following~\cite{BecVex05}, we rewrite the term $\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})$ using the Fundamental Theorem of Calculus,
%
\begin{equation}
\label{eq:Mantiderivative}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \int\limits_{0}^{1} \mathcal{M}'_{HF}\left(\Psi_{HF} + se\right)\left(e\right) \, \textrm{d}s.
\end{equation}
%
Applying the trapezoidal rule to the integral in~\Cref{eq:Mantiderivative} gives,
%
\begin{equation}
\label{eq:Mprimetrapezoid}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{HF})(e) + \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(e) + \mathcal{R}\textrm{.}
\end{equation}
%
where the remainder $\mathcal{R}$ is given as,
%
\begin{equation}
\label{eq:trapezoidremainder}
\mathcal{R} = \frac{1}{2} \int\limits_{0}^{1} \mathcal{M}'''_{HF}\left(\Psi_{LF} + se\right)(e,e,e) \, s(s-1) \, \textrm{d}s,
\end{equation}
%
which is finite due to the third-order differentiability assumptions. We further note that $\mathcal{M}'_{HF}(\Psi_{HF})(e)$ vanishes since $\Psi_{HF}$ is a stationary point of $\mathcal{M}_{HF}$, and~\Cref{eq:Mprimetrapezoid} can be simplified to,
%
\begin{equation}
\label{eq:beckvex}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{R}\textrm{.}
\end{equation}
%
Combining \Cref{eq:repIwithM} and \Cref{eq:beckvex} we obtain,
%
\begin{multline}
\label{eq:preadj}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}\textrm{,}
\end{multline}
which completes the proof.
\end{proof}
%
